Sequential rank agreement methods for comparison of lists

### 2015-08-07 latex/paper.Rnw
Fixing a comment
Answering some comments

Each spectrum consists of $49642$ samples over a mass-to-charge range
of $800$ to $20000$ which we down-sample on an equidistant grid of
5000 points by linear interpolation. After this dimension reduction
the mass spectrum is represented by 5000 predictors. We preprocess
%% ANDREAS: det er uklar om preprocessing forgaar i de oprindelig spectra eller i de down-samplede 
each spectrum by first removing the slow-varying baseline intensity
with the SNIP algorithm \citep{ryan1988snip} followed by a
normalization with respect to the total ion count. We then standardize
the 5000 predictors to have column-wise zero mean and unit variance in
each data set.

@@

All four methods depend on tuning parameters. We select these
separately for each sub-sample by a 20 times repeated 5-fold
cross-validation.
%% ANDREAS: hvilke tuning parameters har du optimeret for Random forest og PLS-DA?
The implementation of LASSO and Ridge regression in the
\texttt{glmnet} package offers three different cross-validated
optimization criteria for the penalty parameter: total deviance,
classification accuracy and area under ROC. We apply all three
criteria to our data to investigate their effect on the agreement.
Note also that the LASSO models produce censored lists depending on
the value of the penalty parameter. Under the deviance optimization
criterion the median number of non-zero coefficients was 33 (range 18
to 52). In order to calculate the sequential rank agreement for the
LASSO we therefore applied our proposed Algorithm \ref{sra-algorithm}
with $B = 1000$ and $P=5000$.

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=.5\textwidth]{pics/riskAgreementPlot}%
\includegraphics[width=.5\textwidth]{pics/predictorAgreementPlot}
\end{center}
%% ANDREAS: kan du aendre titel for begge grafer:
%% "Risk probabilities" -> "Predicted risks of malignant tumor in 113 patients (DACOVA)"
%% "Predictor importance" -> "Discrimination importance of 5000 predictors"
\caption{Left panel: Sequential rank agreement of 1000 rankings of the
  predicted risks of malignant tumor. For each method the different
  rankings were obtained by first training models in 1000 random
  sub-samples of the MALOVA data and then predicting the risk of
  malignant tumor in the 113 MALOVA patients. Right panel: Sequential
  rank agreement of 1000 rankings of the 5000 predictors. The rankings
  were obtained from the same 1000 trained models.}
 \label{fig:app1}
\end{figure}

@@

A not expected shape of the agreement curves is seen for the Ridge
models for all three tuning criteria. They all show higher
disagreement in the top of the lists followed by a decrease. The
reason behind this behavior is rather subtle. Looking at the
distribution of the regression coefficients we see that a large
proportion of them are numerically very close to zero and have almost
equal absolute value. This is a general feature of the Ridge models in
this data set and seen for all the 1000 trained models.
%% ANDREAS: er det sidste ok?
This implies that when predictors are ranked according to the
magnitude of their coefficients, their actual order becomes more
uncertain and more close to a random permutation. This problem can be
alleviated by artificially censoring all predictors with absolute
coefficient values below a given threshold. For the Rigde models tuned
with the deviance criterion, Figure \ref{fig:app2} (left) shows the
the sequential rank agreement where for each of the 1000 trained
models predictors were artificially censored when their absolute
coefficient value was lower than the $0.1\%$ quantile of the 5000
absolute coefficient values. The corresponding curve from Figure
\ref{fig:app1} (left panel) is shown for comparison.
%% ANDREAS: godkender du min aendring af de sidste 2 saetninger?
Thus, in our application the median number of censored predictors
across the 1000 sub-samples was 5 (range 1 to 11). Even though the
number of censored predictors is very small compared to the total
number of predictors, the effect on the sequential rank agreement is
substantial and with the artificial censoring the shape of the curves
is as expected, starting low and then increasing. Also, the effect of
the censoring caused by the sparsity of the LASSO is clearly visible
in the agreement curves.
%% ANDREAS: man kan desvaerre ikke se hvor LASSO agreement forsvinder
%% (ved list-depth 5?) og om den er laver for list depth 1-5 end rigde
%% eller om den er kun laver for list depth 1-2. maaske kan du
%% beskrive dette i teksten?

@@

\section{Discussion}

In this article we address the problem of comparing ranked lists of
the same items. Our proposed method can handle both the situation
where the underlying data to generate the ranked lists are available
and the situation where the only available data is the actual ranked
lists. In addition, censored ranked lists where only the ranks of the
top $k$ ranked items are known can be accommodated as well. The
proposed agreement measure can be interpreted as the average distance
between an item's rank and the average rank assigned to that item
across lists. Thus the measure determines how well the lists agree on
the ranks of a specific item.

>>>>

Each spectrum consists of $49642$ samples over a range of mass-to-charge ratios 
between $800$ to $20000$ Dalton which we downsample on an equidistant grid of
5000 points by linear interpolation. We then preprocess the downsampled 
spectra individually by first removing the slow-varying baseline intensity
with the SNIP algorithm \citep{ryan1988snip} followed by a
normalization with respect to the total ion count. We then standardize
the 5000 predictors to have column-wise zero mean and unit variance in
each data set.

@@

All four methods depend on a tuning parameter. We select these
separately for each sub-sample by a 20 times repeated 5-fold
cross-validation procedure. For Lasso and Ridge regression the tuning parameter
is the degree of penalization, for PLS-DA it is the number of components 
(the dimensionality of the subspace), and for Random Forest it is the number
 of predictors randomly sampled at each split.

The implementation of LASSO and Ridge regression in the
\texttt{glmnet} package offers three different cross-validated
optimization criteria for the penalty parameter: total deviance,
classification accuracy and area under ROC. We apply all three
criteria to our data to investigate their effect on the agreement.
Note also that the LASSO models produce censored lists depending on
the value of the penalty parameter. Under the deviance optimization
criterion the median number of non-zero coefficients was 33 (range 18
to 52). In order to calculate the sequential rank agreement for the
LASSO we therefore applied our proposed Algorithm \ref{sra-algorithm}
with $B = 1000$ and $P=5000$.

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=.5\textwidth]{pics/riskAgreementPlot}%
\includegraphics[width=.5\textwidth]{pics/predictorAgreementPlot}
\end{center}
\caption{Left panel: Sequential rank agreement of 1000 rankings of the
  predicted risks of malignant tumor. For each method the different
  rankings were obtained by first training models in 1000 random
  sub-samples of the MALOVA data and then predicting the risk of
  malignant tumor in the 113 MALOVA patients. Right panel: Sequential
  rank agreement of 1000 rankings of the 5000 predictors. The rankings
  were obtained from the same 1000 trained models.}
 \label{fig:app1}
\end{figure}

@@

A not expected shape of the agreement curves is seen for the Ridge
models for all three tuning criteria. They all show higher
disagreement in the top of the lists followed by a decrease. The
reason behind this behavior is rather subtle. Looking at the
distribution of the regression coefficients we see that a large
proportion of them are numerically very close to zero and have almost
equal absolute value. This is a general feature of the Ridge models in
this data set and seen for all the 1000 trained models.
%% ANDREAS: er det sidste ok?
This implies that when predictors are ranked according to the
magnitude of their coefficients, their actual order becomes more
uncertain and more close to a random permutation. This problem can be
alleviated by artificially censoring all predictors with absolute
coefficient values below a given threshold. For the Ridge models tuned
with the deviance criterion, Figure \ref{fig:app2} (left) shows the
the sequential rank agreement where for each of the 1000 trained
models predictors were artificially censored when their absolute
coefficient value was lower than the $0.1\%$ quantile of the 5000
absolute coefficient values. The corresponding curve from Figure
\ref{fig:app1} (left panel) is shown for comparison.
%% ANDREAS: godkender du min aendring af de sidste 2 saetninger?
Thus, in our application the median number of censored predictors
across the 1000 sub-samples was 5 (range 1 to 11). Even though the
number of censored predictors is very small compared to the total
number of predictors, the effect on the sequential rank agreement is
substantial and with the artificial censoring the shape of the curves
is as expected, starting low and then increasing. Also, the effect of
the censoring caused by the sparsity of the LASSO is clearly visible
in the agreement curves.
%% ANDREAS: man kan desvaerre ikke se hvor LASSO agreement forsvinder
%% (ved list-depth 5?) og om den er laver for list depth 1-5 end rigde
%% eller om den er kun laver for list depth 1-2. maaske kan du
%% beskrive dette i teksten?

@@

\section{Discussion}
In this article we address the problem of comparing ranked lists of
the same items. Our proposed method can handle both the situation
where the underlying data to generate the ranked lists are available
and the situation where the only available data is the actual ranked
lists. In addition, censored ranked lists where only the ranks of the
top $k$ ranked items are known can be accommodated as well. The
proposed agreement measure can be interpreted as the average distance
between an item's rank and the average rank assigned to that item
across lists. Thus the measure determines how well the lists agree on
the ranks of a specific item.
